Effectiveness of private sector malaria control at a large sugar

### 2017-12-02 inst/rmd/paper.Rmd
big updates

We sought to understand the effect of IRS on individual workers' likelihood of absence from work as well as their likelihood of clinical malaria. To estimate this effect, we estimated separate models for absence and illness. We employed interrupted time series [@Lopez_Bernal_2016] and a linear probability approach using the following econometric model.

$$
\hat{Y} = \beta_{0} +  \beta_{1} \text{Season} + (\beta_2{IRS}*\beta_3{IRS_t}) + ... 
$$

$\hat{Y}$ is the rate of (illness/absence). $\beta_{1}$ represents the clinical malaria incidence at that time in the entire district of Manhiça.  Our demographic confounders (represented by $...$) are sex, age, and worker department (field, factory, or administrative). Our intervention was not a simple yes/no, but rather the product of whether the residence of the worker in question was treated in the last year, and, if so, the time since treatment (represented above as the interaction term, where where $_t$ represents time elapsed since commencement of the most recent IRS campaign). We considered using weather data as a right-hand side variable for our models, but found that district-level clinical malaria incidence was a far better predictor of both absence and clinical malaria infection than precipitation and temperature, the former apparently capturing weather's significant effects. Accordingly, we use clinical malaria incidence as a proxy for the severity of malarial activity, and exclude weather variables from our modeling.

Worker characteristics, illness and absenteeism data, along with IRS activity data, were systematically stored, collected, and used at the individual level by Maragra, and therefore of generally high quality. Because cost data was less systematically collected by Maragra, and because many costs could not be precisely quantified due to the abundance of in-kind and cross-departmental expenditures, we had to rely on rough estimations based on a mix of interviews, reciepts, and interpolations. Since our program cost data is not as reliable as our worker characteristic and outcome data, we were conservative in our estimates, and generally tried to err on the side of program activities and materials costing _more_ than what was reported, when doubt was aired. Cost data consisted of three types: (i) wages of malaria control employees, (ii) transporation and vehicle costs for IRS teams, and (iii) acquisition costs of purchasing IRS chemicals for fumigation (ACT and DDT). We define the IRS program as "profitable" if return on investment is greater than 100%, ie if the savings associated with the estimated effect of IRS is greater than the costs of the program's administration.

@@

```{r, fig.cap = 'Clinical malaria (district of Manhiça), all-cause absenteeism among Maragra workers, sick absenteeism among Maragra workers, estimated rainfall, positive cases at company clinic, and test positivity rate at company clinic', fig.height=5, fig.width=8}
# BES incidence
g1 <- 
  ggplot(data = bes,
       aes(x = date,
           y = p)) +
  geom_line(alpha = 0.6) +
  theme_maragra(base_size = 9) +
  labs(x = 'Date',
       y = 'Incidence',
       title = 'A: Annualized district clinical\nmalaria incidence per 1,000')

@@

The below chart shows the effect of IRS. All workers. Reference group is "Never" (those who never got IRS).

```{r, fig.height = 3.5, fig.width = 4, fig.cap= 'Effect of IRS on absenteeism'}
ggplot(data = fit_tidy %>%
         filter(!term %in% c('(Intercept)', 'incidence')) %>%
         filter(grepl('seasonhigh', term)) %>%
         filter(term != 'seasonhigh') %>%
         mutate(term = gsub('seasonhigh:days_since', '', term)) %>%
         mutate(term = factor(term, levels = unique(c('Never',
                                               'Before',
                                               sort(unique(term)))))),
       aes(x = term,
           y = estimate)) +
  geom_bar(stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_maragra() +
  labs(x = 'Time since IRS',
       y = 'Coefficient')
```

>>>>

Worker characteristics, illness and absenteeism data, along with IRS activity data, were systematically stored, collected, and used at the individual level by Maragra, and therefore of generally high quality. 
We sought to understand the effect of IRS on individual workers' likelihood of absence from work as well as their likelihood of clinical malaria. To estimate this effect, we estimated separate models for absence and illness. We employed interrupted time series [@Lopez_Bernal_2016] and a linear probability approach using the following econometric model.

$$
\hat{Y_{it}} = \beta_{0} +  (\beta_{1}) (\text{Season}_{t}) + (\beta_2{IRS}*\beta_3{IRS_t}) + ... + \epsilon
$$

$\hat{Y}$ is the rate of absence. $\beta_{1}$ represents the clinical malaria incidence at that time in the entire district of Manhiça.  Our demographic confounders (represented by $...$) are sex, age, and worker department (field, factory, or administrative). Our intervention was not a simple yes/no, but rather the product of whether the residence of the worker in question was treated in the last year, and, if so, the time since treatment (represented above as the interaction term, where where $_t$ represents time elapsed since commencement of the most recent IRS campaign). We define the malaria season as any time during which the clinical incidence of malaria in the district of Manhiça was at or greater than the median clinical incidence of malaria for the entire study period. These weeks are flagged as red in Figure 1, Panel A.  By using clinical incidence of the area of residence of the workers (as opposed to more typical proxies for malaria risk, such as only rainy vs. non rainy season), our seasonality estimate is a closer approximation of true malaria risk, incorporating lagged effects such as the incubation period of the parasite, as well as any inherent non-linear effects of weather. In addition, we adjust for daily precipitation

Because cost data was less systematically collected by Maragra, and because many costs could not be precisely quantified due to the abundance of in-kind and cross-departmental expenditures, we had to rely on rough estimations based on a mix of interviews, reciepts, and interpolations. Since our program cost data is not as reliable as our worker characteristic and outcome data, we were conservative in our estimates, and generally tried to err on the side of program activities and materials costing _more_ than what was reported, when doubt was aired. Cost data consisted of three types: (i) wages of malaria control employees, (ii) transporation and vehicle costs for IRS teams, and (iii) acquisition costs of purchasing IRS chemicals for fumigation (ACT and DDT). We define the IRS program as "profitable" if return on investment is greater than 100%, ie if the savings associated with the estimated effect of IRS is greater than the costs of the program's administration.

@@

```{r, results='asis'}
calculate <- function(var){
  out <- 
  model_data %>%
    mutate(year = format(date, '%Y')) %>%
    mutate(maragra_fabrica = ifelse(maragra_fabrica,
                                    'On site',
                                    'Off site'),
           rainy = ifelse(rainy,
                          'Rainy',
                          'Dry')) %>%
    filter(!is.na(rainy)) %>%
  group_by_(var, 'year') %>%
  summarise(absences = length(which(absent)),
            eligibles = length(absent),
            workers = length(unique(oracle_number))) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100) 
  names(out)[1] <- 'variable'
  out$val <- paste0(round(out$absenteeism_rate, digits = 2),
                    '%')#,
                    # ')',
                    # '\n',
                    # ' (workers:', out$workers,
                    # ')')
  out <- 
    out %>%
    dplyr::select(variable, year, val)

  out <- out %>%
    spread(key = year,
           value = val)
  out$variable <- as.character(out$variable)
  new_var <- ifelse(var == 'maragra_fabrica',
                    'Residence',
                    ifelse(var == 'rainy',
                           'Precipitation',
                           var))
  out$variable[1] <- paste0(Hmisc::capitalize(new_var), ': ', out$variable[1])
  return(out)
}
out_list <- list()
vars <- c('season', 
           'department',
           'sex',
           'maragra_fabrica',
           'rainy')
for(i in 1:length(vars)){
  x <- calculate(var = vars[i])
  out_list[[i]] <- x
}
ab_table <- bind_rows(out_list)
names(ab_table)[1] <- ''
print(xtable::xtable(ab_table,
                     align = c('r' ,'r', rep('l', 4))),
      include.rownames = FALSE)
```

```{r, fig.cap = 'Clinical malaria (district of Manhiça), all-cause absenteeism among Maragra workers, sick absenteeism among Maragra workers, estimated rainfall, positive cases at company clinic, and test positivity rate at company clinic', fig.height=5, fig.width=8}
# BES incidence
g1 <- 
  ggplot(data = bes %>%
           mutate(season = ifelse(p >= median(p), 'High season',
                                  'Low season')) %>%
           filter(weekdays(date) == 'Monday'),
       aes(x = date,
           y = p)) +
  geom_point(alpha = 0.6,
             aes(color = season),
             size = 0.2) +
  geom_line(alpha = 0.1) +
  theme_maragra(base_size = 9) +
  labs(x = 'Date',
       y = 'Incidence',
       title = 'A: Annualized district clinical\nmalaria incidence per 1,000') +
  scale_color_manual(name = '',
                     values = c('red', 'black')) + 
  guides(color = guide_legend(ncol = 1)) +
  theme(legend.position = c(.8, .8),
        legend.text = element_text(size = 5),
        legend.background = element_rect(fill=alpha('white', 0)))

@@

```{r, fig.height = 3.5, fig.width = 6, fig.cap = 'Effect of IRS on absenteeism by season'}
# Predict for visualization
dummy <- expand.grid(season = sort(unique(model_data$season)),
                     months_since = sort(unique(model_data$months_since)))
predictions <- predict(fit, dummy, se.fit = TRUE, interval = 'confidence')
dummy$predicted <- predictions$fit[,1]
dummy$lwr <- predictions$fit[,2]
dummy$upr <- predictions$fit[,3]
ggplot(data = dummy %>%
         filter(months_since != 'Never') %>%
         mutate(lwr = lwr * 100,
                upr = upr * 100,
                predicted = predicted * 100) %>%
         mutate(season = paste0(Hmisc::capitalize(as.character(season)), ' season')),
       aes(x = months_since,
           y = predicted,
           group = 1)) +
  geom_ribbon(aes(x = months_since,
                  ymin = lwr,
                  ymax = upr),
              fill = 'lightblue',
              alpha = 0.6) +
  # geom_line(color = 'blue', alpha = 0.5) +
  geom_point(color = 'blue', alpha = 0.5) + 
  geom_smooth(se = FALSE,
              color = 'black',
              alpha = 0.5,
              span = 1) +
  facet_wrap(~season) +
  theme_maragra() +
  theme(axis.text.x = element_text(angle = 90, size = 8)) +
  labs(x = 'Months since IRS',
       y = 'Absenteeism (%)')

```
